Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-03-09 paper/paper.tex
update

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be quantified locally,
including the independent set problem, the cutting problem, the 3-coloring problem, the 3-satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
\blue{need more diversed applications not limited to IS}
In combinatorial optimization, the most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. people believe there is unlikely a polynomial time algorithm for solving them.
In this paper, we study the property computing that can be harder but also more useful for people to understand a problem.
The properties include finding the maximum/minimum size, counting sets of a give size, enumerating all such sets or direct sampling them when they are not enumeratable.
For weighted version, it also includes finding sets with the largest/smallest $k$ sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

For avoid the discussion being too diversed, we limit the target problem to the independent set problem,
and discuss other combinatorial optimization problems in \App{app:otherproblems}.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$. 
One of the paradigmatic independent set \textit{properties} % that many computational complexity theorists are most interested in 
is the maximum independent set (MIS) and the size of the MIS, $\alpha(G) \equiv \max_{I}|I|$.

On the other hand, some other properties, such as the number and enumeration of independent sets or maximal independent sets of a given size, are equally interesting, if not more important, in many practical applications. These other properties can be found in physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
they can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007}, to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}. However, there is a lack of general and versatile tool to compute these different properties of independent sets.

In this paper, we introduce a generic programming tensor-network framework to compute a number of different independent set properties.
We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
It is different from the standard tensor network methods in that the tensor elements are not restricted to standard number types such as floating point numbers and integers.
Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
The complexity of our tensor-network based algorithms is similar to that of dynamic programming algorithms~\cite{Courcelle1990, Fomin2013}, which have a complexity scaling exponentially with the treewidth, ${\rm tw}(G)$, of the graph;
for sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.
We benchmark our algorithms in \Sec{sec:benchmark} by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
The high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021},
where efficient heuristic algorithms have been developed to optimize the contraction order for tensor networks up to thousands of vertices.
For cases where the tensors are too large to fit into a GPU, slicing techniques can be used to further reduce the space consumption.
Lastly, we provide a few examples and demonstrate the versatility of our tool by computing the entropy constant for some hardcore lattice gases on 2D square lattices and analyzing the presence of overlap gap property on three regular graphs.
Our method can also be used to find ``maximal'' independent set properties; 
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

\section{Tensor network representation of independent sets} \label{sec:indpoly}
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

\section{Independence polynomial}

@@

\section{Maximum independent sets and its counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of a given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\section{Enumerating and sampling independent sets}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

\section{Weighted graphs}
In the previous discussion, we have limited ourselves to unweighted graphs.
For integer weighted graphs, all results still holds,
while for general weighted graphs, the independence polynomial is not computable anymore.
For general weighted graphs, we are more interested to know the total weights of maximum $k$ independent set sizes and their configurations.
Hence we define the extended tropical numbers as the following

@@

\section{Benchmarks and case studies}\label{sec:benchmark}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

\subsection{Example case studies} In this section, we give a few examples where the different properties of independence sets are used.

\subsubsection{Number of independent sets and entropy constant for some hardcore lattice gases}\label{sec:entropy}
We compute the counting of all independent sets for graphs shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The types of graphs include: the square grid graphs (Figure~\ref{fig:lattices}(a)); the square grid graphs with a filling factor $p=0.8$, which means $\lfloor pL^{2} \rceil$ square grids are occupied with vertices  (Figure~\ref{fig:lattices}(b));
the King's graphs  (Figure~\ref{fig:lattices}(c)); the King's graphs with a filling factor $p = 0.8$  (Figure~\ref{fig:lattices}(d)).
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequence for $L=38$ and $L=39$, which, to the best of our knowledge, is not known before.
In the computation, we used the finite-field algebra for contracting arbitrarily high precision integer tensor networks.

@@

\subsubsection{The overlap gap property}
With the tool to enumerate configurations, one can try to understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs and the geometry of the solution space. One of the known barriers for finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}. Intuitively, if the overlap gap property is present, it means every two large independent sets either have a significant intersection or very small intersection; it implies that large independent sets are clustered together. This clustering property has been used to prove the limitations of local algorithms in finding the MISs~\cite{Gamarnik2013, Gamarnik2019}. Here, we use the computed configurations of large independent sets to investigate the presence or absence of the overlap gap property for some given graphs. In particular, in \Fig{fig:hamming}, we enumerated all ISs of size $\geq\alpha(G)-1$ for two random three regular graph instances of size 100 and show the pairwise Hamming distance statistics for the enumerated configurations. In \Fig{fig:hamming}(a), we observed the multiple peak structure for the pairwise Hamming distance for one of the graphs. This indicates the presence of the overlap gap property and disconnected clusters exist in the configuration space for large independent sets. We expect our numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of pairwise Hamming distances between every two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}

>>>>

\section{Introduction}
In graph theory and combinatorial optimization, there is an important class of problems that can be quantified locally,
including the independent set problem, the cutting problem, the 3-coloring problem, the 3-satisfiability problem,
the maximal clique problem, and the vertex cover problem et al ~\cite{Moore2011}.
There is a widespread interest of these problem because they have wide range of applications in scheduling, logistics,
wireless networks and telecommunication, and computer vision, etc.~\cite{Butenko2003, Wu2015}.
\blue{need more diversed applications not limited to IS}

The most studied version of the above problems is finding the maximum/minimum set size and one of the solution with that size satisfying the contraints,
while many of them belong to the hardness category NP-hard~\cite{Hastad1996}, i.e. there is unlikely a polynomial time algorithm for solving them.
In this paper, we focus on the less studied \textit{property} computation of these problems which can be harder but is crutial towards understanding a problem.
The property computation includes finding the maximum/minimum set size, counting sets of a give size, enumerating all such sets or direct sampling them when they are not enumeratable.
For the weighted version, it also includes finding largest/smallest $k$ sets and their sizes.
%are many interesting and hard computational problems concerning various properties of independent sets.

We provide an abstract algebra view for the computation of all these properties and proposed a unified framework,
generic tensor-network, to solve them all.
%We map the computation of these properties into generic tensor network contraction with specially designed tensor element algebra.
Tensor network is also known as the sum product network or einsum in different fields,
where its tensor element types are typically restricted to standard number types such as real numbers and complex numbers.
While generic tensor network is a generalization of which by extending the tensor element type to a generic one being a commutative semiring.
%Traditional algorithms for finding the MIS size include the branching algorithms~\cite{Tarjan1977, Robson1986} and dynamic programming based algorithms~\cite{Courcelle1990, Fomin2013}.
Given a good contraction order, contracting a tensor network in general has time complexity $2^{{\rm tw}(G)}$, where ${\rm tw}(G)$ is the treewidth of graph $G$.
Recent progress in simulating quantum circuits with random tensor networks~\cite{Gray2021, Pan2021, Kalachev2021} enables us to find the optimal contraction order of a large scale tensor network in reasonable time.
Although this time complexity is just on par with the dynamic programming algorithms~\cite{Courcelle1990, Fomin2013} for solving the maximum size and one of the best set, tensor network has richer algebraic structure, and this structure is crutial for computing properties.
%For sparse graphs, the treewidth is usually much smaller than the number of vertices~\cite{Fomin2006}. Our algorithms, on the other hand, are much more versatile than traditional methods and can be used to compute many other properties than just the MIS size.

To avoid the discussion being overly diversed, we limit the target problem to the independent set problem,
and discuss other combinatorial optimization problems in \App{app:otherproblems}.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$,
and we denote the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$. 
%The independent set problem is related to physics applications such as in the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
%in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018, Ebadi2022};
%Property computation can, for example, be used to understand phase transitions~\cite{Dyre2016, Fernandes2007},
%to identify harder graphs in an ensemble of graphs~\cite{Ebadi2022} and to analyze the presence of the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} related to the clustering phenomenon of large independent sets.
Meanwhile, finding the number of independent sets of a given size is equivalent to computing the coefficients of the independence polynomial~\cite{Harvey2018,Ferrin2014}.
which is a useful graph characteristic related to, for example, the partition functions~\cite{Lee1952,Yang1952} and Euler characteristics of the independence complex~\cite{Bousquet2008, Levit2009}.
%However, there is a lack of general and versatile tool to compute these different properties of independent sets.

The paper is organized as follows.
We first introduce the basic concepts of tensor networks and generic programming in \Sec{sec:tn} and \Sec{sec:generic}.
Then we show how to reduce an independent set problem to a tensor network contraction problem \Sec{sec:tnmap}
and show how to engineer the element types in \Sec{sec:counting},\Sec{sec:enumeration} and \Sec{sec:weighted} to count independent sets at a given size, enumerate/sample independent sets at a given size and find largest set sizes in weighted graphs.
% by computing a number of independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance.
Lastly, we benchmark our algorithms in \Sec{sec:benchmark} and provide a few examples in \Sec{sec:examples} to demonstrate the versatility of our tool.
We show how our method can help computing the entropy constant for some hardcore lattice gases on 2D square lattices
and analyzing the presence of overlap gap property on three regular graphs.
%Our method can also be used to find ``maximal'' independent set properties; 
%a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.

\section{Tensor networks}\label{sec:tn}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
Let $G=(V,E)$ be a weighted graph with each vertex $i\in V$ being associated with a weight $w_i$.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i, w_i)$ as
\begin{equation}
    W(x_i, w_i) = \left(\begin{matrix}
        1 \\
        x_i^{w_i}
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i, w_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i, w_i)_1=x_i^{w_i}$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i, w_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.

@@

\section{Independence polynomial}\label{sec:indpoly}

@@

\section{Maximum independent sets and its counting}\label{sec:counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of a given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with a new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
If one is only interested in finding the MIS size, one can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm T} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm T} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\section{Enumerating and sampling independent sets}\label{sec:enumeration}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

@@

\section{Weighted graphs}\label{sec:weighted}
In the previous discussion, we have limited ourselves to unweighted graphs.
For integer weighted graphs, all results still holds,
while for general weighted graphs, the independence polynomial is not computable anymore.
For general weighted graphs, we are more interested to know the total weights of maximum $k$ independent set sizes and their configurations.
Hence we define the extended tropical numbers as the following

@@

\section{Performance benchmarks}\label{sec:benchmark}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small treewidth that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

\section{Example case studies} \label{sec:examples}
In this section, we give a few examples where the different properties of independence sets are used.

\subsection{Number of independent sets and entropy constant for some hardcore lattice gases}\label{sec:entropy}
We compute the counting of all independent sets for graphs shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The types of graphs include: the square grid graphs (Figure~\ref{fig:lattices}(a)); the square grid graphs with a filling factor $p=0.8$, which means $\lfloor pL^{2} \rceil$ square grids are occupied with vertices  (Figure~\ref{fig:lattices}(b));
the King's graphs  (Figure~\ref{fig:lattices}(c)); the King's graphs with a filling factor $p = 0.8$  (Figure~\ref{fig:lattices}(d)).
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequence for $L=38$ and $L=39$, which, to the best of our knowledge, is not known before.
In the computation, we used the finite-field algebra for contracting arbitrarily high precision integer tensor networks.

@@

\subsection{The overlap gap property}
With the tool to enumerate configurations, one can try to understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs and the geometry of the solution space. One of the known barriers for finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}. Intuitively, if the overlap gap property is present, it means every two large independent sets either have a significant intersection or very small intersection; it implies that large independent sets are clustered together. This clustering property has been used to prove the limitations of local algorithms in finding the MISs~\cite{Gamarnik2013, Gamarnik2019}. Here, we use the computed configurations of large independent sets to investigate the presence or absence of the overlap gap property for some given graphs. In particular, in \Fig{fig:hamming}, we enumerated all ISs of size $\geq\alpha(G)-1$ for two random three regular graph instances of size 100 and show the pairwise Hamming distance statistics for the enumerated configurations. In \Fig{fig:hamming}(a), we observed the multiple peak structure for the pairwise Hamming distance for one of the graphs. This indicates the presence of the overlap gap property and disconnected clusters exist in the configuration space for large independent sets. We expect our numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of pairwise Hamming distances between every two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}
